\documentclass[12pt]{article}

\input{../../../hw_packages.tex}

\title{Homework 12: Phys 7320 (Spring 2022)}
\due{April 20, 2022}

\begin{document}
\maketitle
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{problem}{12.1}[The Lagrangian for a charged particle]
The Lagrangian for a charged particle of mass $m$ and charge $e$ with position
$\vb{r}$ and velocity $\vb{u}\equiv d\vb{r}/dt$ moving in scalar and vector
potentials $\Phi$ and $\vb{A}$ is
\begin{equation}
    \LL=-mc^2\sqrt{1-\frac{u^2}{c^2}}-e\Phi+\frac{e}{c}\vb{u}\vdot\vb{A}.
\end{equation}

(a) Show that the Euler-Lagrange equations for this Lagrangian indeed give rise
to the Lorentz force law. \textit{Hint}: I suggest using index notation, and
remember the potentials $\Phi,\vb{A}$ can depend on both $\vb{r}$ and $t$, which
as far as the particle is concerned means they depend on time in two ways:
$\Phi(\vb{r}(t),t)$ and $\vb{A}(\vb{r}(t),t)$.

(b) Go though the steps (12.13)-(12.17) in Jackson to derive the Hamiltonian
\begin{equation}
    \HH=\sqrt{(c\vb{P}-e\vb{A})^2+m^2c^4}+e\Phi. 
\end{equation}
Along the way derive the canonical/conjugate momentum $\vb{P}$ (which is
different from the familiar ``kinematic momentum'' $\vb{p}=\gamma m\vb{u}$), and
invert it to find an expression for $\vb{u}$ in terms of $\vb{P}$ and $\vb{A}$.

If you want extra practice, you can show that Hamilton's equations for this
Hamiltonian also give rise to the Lorentz force law, with some steps similar to
part (a).
\begin{solution}
(a) First, we write $\LL$ in Einstein notation with $j\in\qty{x,y,z}$
\begin{equation}
    \LL=-mc^2\sqrt{1-\frac{u_j^2}{c^2}}-e\Phi+\frac{e}{c}u_jA_j. 
\end{equation}
Then it is clear that
\begin{equation}\label{p1a:P}
    \frac{\partial\LL}{\partial\dot{r}_i}
    =\frac{\partial\LL}{\partial u_i}
    =\frac{mu_i}{\sqrt{1-u_j^2/c^2}}+\frac{e}{c}A_i
    \Rightarrow
    \frac{\partial\LL}{\partial\dot{\vb{r}}}
    =\gamma m\vb{u}+\frac{e}{c}\vb{A}
    =\vb{p}+\frac{e}{c}\vb{A}.
\end{equation}
Thus, the LHS of Euler-Lagrange equation is
\begin{equation}\label{p1a:LHS}
    \frac{d}{dt}\frac{\partial\LL}{\partial\dot{\vb{r}}}
    =\frac{\partial\vb{p}}{\partial t}
    +\frac{e}{c}\frac{\partial\vb{A}}{\partial t}
    +\frac{e}{c}(\vb{u}\vdot\grad)\vb{A}.
\end{equation}
Now, for the RHS of Euler-Lagrange equation, we can write
\begin{equation}\label{p1a:RHS}
    \grad\LL=-e\grad\Phi+\frac{e}{c}\grad(\vb{u}\vdot\vb{A}) 
    =-e\grad\Phi+\frac{e}{c}\qty[(\vb{u}\vdot\grad)\vb{A}+\vb{u}\times(\curl{\vb{A}})],
\end{equation}
where we have used the vector calculus identity
$\grad(\vb{A}\vdot\vb{B})=(\vb{A}\vdot\grad)\vb{B}+(\vb{B}\vdot\grad)\vb{A}+\vb{A}\times(\curl{\vb{B}})+\vb{B}\times(\curl{\vb{A}})$.
Now, combining \eqref{p1a:LHS} and \eqref{p1a:RHS}, we can write
\begin{equation}
    \frac{\partial\vb{p}}{\partial t}
    =e\qty(-\grad\Phi-\frac1c\frac{\partial\vb{A}}{\partial t})
    +e\frac{\vb{u}}{c}\times(\curl{\vb{A}})
    =e\qty(\vb{E}+\frac{\vb{u}}{c}\times\vb{B}),
\end{equation}
which is Lorentz force law.

(b) From \eqref{p1a:P}, we have already found the canonical momentum
\begin{equation}
    \vb{P}=\gamma m\vb{u}+\frac{e}{c}\vb{A}. 
\end{equation}
Thus,
\begin{equation}
    \vb{u}=\frac{c\vb{P}-e\vb{A}}{\gamma mc}
    \Rightarrow u^2=\frac{(c\vb{P}-e\vb{A})^2}{\gamma^2m^2c^2}.
\end{equation}
But also, the Lorentz factor by definition is
\begin{equation}
    \gamma^2=\frac{1}{1-u^2/c^2}=\frac{\gamma^2m^2c^4}{\gamma^2m^2c^4-(c\vb{P}-e\vb{A})^2}.
\end{equation}
Inverting, we get
\begin{equation}
    \gamma=\sqrt{1+\qty(\frac{\vb{P}}{mc}-\frac{e\vb{A}}{mc^2})^2}. 
\end{equation}
Thus, $\vb{u}$ in terms of $\vb{P}$ and $\vb{A}$ is
\begin{equation}
    \vb{u}=\frac{c\vb{P}-e\vb{A}}{\sqrt{m^2c^2+\qty(\vb{P}-e\vb{A}/c)^2}}.
\end{equation}
Then, starting from (12.15, Jackson),
\begin{align}
    \HH
    &=\vb{P}\vdot\vb{u}-\LL\notag\\
    &=\frac{\vb{u}}{c}\vdot\qty(c\vb{P}-e\vb{A})+\frac{mc^2}{\gamma}+e\Phi\notag\\
    &=\frac{(c\vb{P}-e\vb{A})^2}{\sqrt{m^2c^4+(c\vb{P}-e\vb{A})^2}}
    +\frac{m^2c^4}{\sqrt{m^2c^4+(c\vb{P}-e\vb{A})^2}}+e\Phi\notag\\
    &=\sqrt{m^2c^4+(c\vb{P}-e\vb{A})^2}+e\Phi.
\end{align}

\end{solution}
\end{problem}
\newpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%    
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{problem}{12.2}[Equivalent Lagrangians]
(a) Use the Principle of Least Action (really the principle of extremal action)
to show that if the Lagrangian $\LL$ is changed by adding the time derivative of
some function of the coordinates and time, then the Euler-Lagrange equations are
unchanged. The new and old Lagrangians are said to be \textit{equivalent}.
Generalize this to a statement about what change to a Lagrangian
\textit{density} $\LL$ leaves the EL equations unchanged.

(b) Show that under a gauge transformation, the Lagrangian for a charged
particle given in the previous problem becomes an equivalent Lagrangian, thus
showing the equations of motion do not change.
\begin{solution}
\end{solution}
\end{problem}
\newpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{problem}{12.3}[$SO(2)$ symmetry of two real scalar fields.]
Consider the dynamics of two (real) scalar fields $\phi_1(\vb{x},t)$ and
$\phi_2(\vb{x},t)$ specified by the Lagrangian density
\begin{equation}
    \LL=\frac12\partial_\mu\phi_1\partial^\mu\phi_1+\frac12\partial_\mu\phi_2\partial^\mu\phi_2-V(\phi_1,\phi_2), 
\end{equation}
where the potential $V$ depends only on the combination $\phi_1^2+\phi_2^2$. In
class, we will study this case with the real scalars combined into a single
compex scalar; here we will leave them as two real scalars. Let's make a
definite choice for the potential:
\begin{equation}
    V(\phi_1,\phi_2)=\frac12m^2\phi_1^2+\frac12m^2\phi_2^2+\frac{\lambda}2\qty(\phi_1^2+\phi_2^2)^2. 
\end{equation}

(a) Calculate the equations of motion (Euler-Lagrange equations) for both
$\phi_1$ and $\phi_2$.

(b) Show that the $SO(2)$ transformation
\begin{equation}
    \mqty(\phi_1\\\phi_2)\mapsto\mqty(\phi_1'\\\phi_2')=\mqty(\cos\alpha&\sin\alpha\\-\sin\alpha&\cos\alpha)\mqty(\phi_1\\\phi_2), 
\end{equation}
where $\alpha$ is a constant, is a symmetry of the Lagrangian. Is this a
rotation in physical space? What space does this ``rotation'' act on?

(c) According to Noether's theorem, the existence of this symmetry means there
is a corresponding conserved current $J^\mu$. Find $J^\mu$ in terms of $\phi_1$
and $\phi_2$ (you may drop an overall constant $\alpha$) and show that it is
conserved, $\partial_\mu J^\mu=0$, when you use the equations of motion.
\begin{solution}
\end{solution}
\end{problem}
\newpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\end{document}
